CODECHECK: An Open Science initiative for the independent execution of computations underlying research articles during peer review to improve reproducibility

### 2021-02-23 codecheck-paper.tex
about to tackle the TOMS bit

\textbf{1. Codecheckers record but don't investigate or fix.} \\
The codechecker follows the author's instructions to run the code. If
instructions are unclear, or if code does not run, the codechecker
tells the author. We believe that the job of the codechecker is not to
fix all these problems but simply to report them and await a fix from
the author.  The level of documentation that is needed for third
parties to reproduce a workflow is hard to get right, and
too often this uncertainty leads researchers to give up and not
document it at all.  The conversation with a codechecker fixes this
problem.

\textbf{2. Communication between humans is key.} \\ 
Some code may
work without any interaction (e.g., CODECHECK certificate \texttt{2020-013}
\cite{cert-2020-013}) but more often there are hidden
dependencies or installations that need adjusting for a particular system.
Allowing the (human) codechecker to communicate
directly and openly with each other should make this process as
constructive as possible. We feel that routing this conversation
(possibly anonymously) through a publisher would introduce delays and
reduce the chances for community building.

\textbf{3. Credit is given to codecheckers.} \\
The value of performing a CODECHECK is comparable to that of a peer
review, and it may also require a similar amount of time. Therefore,
the codechecker's activity should be recorded, ideally in the
published paper.  The public record can be realised by publishing the
certificate in a citable form (i.e.,~with a DOI, as conducted in the
community CODECHECK process), by listing codecheckers on the journal's
website or, ideally, by publishing the checks alongside peer review
activities in public databases.

\textbf{4. Workflows must be auditable.} \\
The codechecker should have sufficient material to validate the
workflow outputs submitted by the
authors. Stark~\cite{stark_before_2018} calls this
``preproducibility'' and the ICERM report \cite{stodden_setting_2013}
defines the level ``Auditable Research'' similarly.  Every community
may establish their own good practices or adapt generic concepts and
practical tools, such as publishing all building blocks of science in
a research compendium (cf.~\url{https://research-compendium.science/})
or ``repro-pack'' \cite{barba_praxis_2018}.
A completed
check means that code could be executed at least once using the
provided instructions, and, therefore, all code and data was given and
could be investigated more deeply or extended in the future.  Ideally,
this is a ``one~click'' step, but achieving this requires particular
skills and a sufficient level of documentation for third
parties. Furthermore, automation may lead to people gaming the system
or reliance on technology, which can often hide important details.
All such aspects can reduce the understandability of the material, so
we estimate our approach to codechecking, done without automation and
with open human communication, to be a simple way to ensure long-term
transparency and usefulness.  We acknowledge that others have argued
in favour of bitwise reproducibility because, in the long run, it can
be automated (e.g., Konrad Hinsen on Twitter,
\url{https://twitter.com/khinsen/status/1242842759733665799}), but
until then we need CODECHECK's approach.

\textbf{5. Open by default and transitional by disposition.} \\
By default, unless there are strong reasons to the contrary
(e.g.,~sensitive data on human subjects), all code and data, both from
author and codechecker, will be made freely available before or when
the certificate is published.  Openness is not required for the paper
itself, to accommodate journals in their transition to sustainable
Open Access models.  The code and data publication should follow
community good practices.  Ultimately we may find that CODECHECKS are
subsumed within peer review.
%% By disposition, a CODECHECK workflow should
%% be designed (a) recognising shortcomings in asserting reproducibility
%% of submissions and (b) with an understanding of the complex situation
%% that led to this. However, establishing a CODECHECK workflow should
%% mirror initiatives to improve education around reproducibility of
%% methods.  The long term goal is for a specific community to reach a
%% state where the level of verification ensured by codecheckers becomes
%% part of peer review and a regular merit of all accepted papers.

@@

We have a curated list of 25 certificates available at
\url{https://codecheck.org.uk/register}.
Table~\ref{tab:register} lists the certificates
along with citations to the original papers.
These CODECHECKs fall into three themes:
(1) classic and current papers from computational neuroscience,
(2) COVID-19 modelling preprints, and
(3) GIScience.

The first theme was the initial set of papers used to explore the
concept of CODECHECK.  The idea was to take very well known historical
articles from a domains of interest to one of the co-authors (SJE).
Our first CODECHECK (certificate number \texttt{2020-001}) was
performed before publication on an article for the journal
\emph{GigaScience}, which visusalized the outputs from a family of
supervised classification algorithms.

The second theme was a response to the COVID-19 pandemic, choosing
modelling papers to predict outcomes. The checks were solicited
through community interaction or by our initiative rather than
requested from journals.  Some certificates were since acknowledged in
the accepted papers
\cite{Davies2020-vj,kucharski_effectiveness_2020}. In particular, we
codechecked the well-known Imperial college model of UK lockdown
procedures from March 2020(certificate \texttt{2020-10}).  Contrary to
opinion in the mainstream media, the model results were reproducible
(CITE NOORDEN).

@@

At the end of each CODECHECK, the codechecker writes a certificate
stating which outputs from the original article, i.e.,~numbers,
figures or tables, could be reproduced.  This certificate is made
openly available so that all readers of the article can see which
elements were reproduced and what limitations or issues were found.
The certificate links to code and data used by the codechecker,
allowing others to build on the work.  The format of the certificates
evolved during the project, as we have learnt to automate different
aspects of the certification, such as inserting figures into
certificate documents and managing metadata of a check.  The metadata
is stored in a machine-readable structured file in YAML, the CODECHECK
configuration file \texttt{codecheck.yml}.
The technical specification of the CODECHECK configuration file is
published at \url{https://codecheck.org.uk/spec/config/latest/}.  The
configuration file enables current and future automation of workflows
and meta-analyses.

Figure~\ref{fig:cert} shows pages 1--4 (of 10) of a typical
certificate \texttt{2020-012} \cite{cert-2020-012}.  The referenced
article \cite{unwin_report_2020} described models of COVID-19 across
the USA.  Figure~\ref{fig:cert}A shows the certificate number and its
DOI, which points to the certificate and any supplemental files on
Zenodo.  The CODECHECK logo is added for recognition and to denote
successful reproduction.  Figure~\ref{fig:cert}B provides the key
metadata as a table generated from \texttt{codecheck.yml}; it names
the paper that was checked (title, DOI), the authors, the codechecker,
when the check was performed, and in which repository related material
is accessible.  Figure~\ref{fig:cert}C shows a textual summary of how
the CODECHECK was performed and any interesting findings.
Figure~\ref{fig:cert}D (page 2 of the certificate) shows the outputs
that were generated based on the MANIFEST of output files in the
CODECHECK.  It shows the file name (Output), the description stating
to which figure/table each file should be compared in the original
paper (Comment), and the file size.  Page 3 of the certificate,
(Figure~\ref{fig:cert}E) gives detailed notes from the codechecker,
here documenting what steps were needed to run the code and that the
code took about 17~hours to complete. Finally, page 4 of the
certificate shows the first output generated by the CODECHECK
(Figure~\ref{fig:cert}F). In this case, the figure matched figure~4 of
\cite{unwin_report_2020}.  The remaining pages of the certificate show
other outputs and the computing environment in which the certificate
itself was created (not shown here).

@@

We use freely available infrastructure, GitHub and Zenodo, to run our
system.  The \texttt{codecheckers} \textbf{GitHub}
organisation at \url{https://github.com/codecheckers} contains
projects for managing the project website, the codecheckers community
and its discussions, code repositories, and the main register of
CODECHECKs. Both the project website \url{https://codecheck.org.uk/}
and the register at \url{https://codecheck.org.uk/register}
are hosted as GitHub pages.
The register database is a single table in CSV format that connects the
certificate identifier with the repository associated with a specific
conducted CODECHECK. Each of these repositories, which currently can
be hosted on GitHub or Open Science Framework (OSF), contains the
CODECHECK metadata file \texttt{codecheck.yml}. The register further
contains a column for the type of check, e.g.,~community, journal, or
conference, and the respective GitHub issue where communications and
assignments around a specific check are organised. No information is
duplicated between the register and the metadata files. The continuous
integration infrastructure of GitHub, GitHub Actions, is used to
automate the register for viewing (HTML) and for integration with
other services (CSV, JSON).  \textbf{Zenodo} is our preferred open repository for
storing certificates. It mints DOIs for deposits and ensures
long-term availability of all digital artefacts related to the
project. The CODECHECK community on Zenodo is available at
\url{https://zenodo.org/communities/codecheck/}. It holds
certificates, the regularly archived register
\cite{codecheck_register_jan2021}, and other material related to
CODECHECK.

@@

These tools and resources demonstrate that a CODECHECK process can be
managed without infrastructure costs on freely available platforms.
The automation should ensure that more checks can be conducted in the
future, though having the process be managed by a paid CODECHECK
editor or further automation, e.g.,~with a bot, to reduce turnaround
time. Our main resource requirements are the \textbf{humans} needed for managing
the project and processes and the codecheckers.  All contributions
currently rely on (partly grant-based) public funding and
volunteering.

>>>>

\textbf{1. Codecheckers record but don't investigate or fix.} \\
The codechecker follows the author's instructions to run the code. If
instructions are unclear, or if code does not run, the codechecker
tells the author. We believe that the job of the codechecker is not to
fix these problems but simply to report them to the author and await a
fix.  The level of documentation required for third parties to
reproduce a workflow is hard to get right, and too often this
uncertainty leads researchers to give up and not document it at all.
The conversation with a codechecker fixes this problem.

\textbf{2. Communication between humans is key.} \\
Some code may work without any interaction, e.g. \cite{cert-2020-013},
but often there are hidden dependencies that need adjusting for a
particular system.  Allowing the codechecker to communicate directly
and openly with the author make this process as constructive as
possible; routing this conversation (possibly anonymously) through a
publisher would introduce delays and inhibit community building.

\textbf{3. Credit is given to codecheckers.} \\
The value of performing a CODECHECK is comparable to that of a peer
review, and it may require a similar amount of time. Therefore, the
codechecker's activity should be recorded, ideally in the published
paper.  The public record can be realised by publishing the
certificate in a citable form (i.e.,~with a DOI), by
listing codecheckers on the journal's website or, ideally, by
publishing the checks alongside peer review activities in public
databases.

\textbf{4. Workflows must be auditable.} \\
The codechecker should have sufficient material to validate the
workflow outputs submitted by the
authors. Stark~\cite{stark_before_2018} calls this
``preproducibility'' and the ICERM report \cite{stodden_setting_2013}
defines the level ``Auditable Research'' similarly.  Communities can
establish their own good practices or adapt generic concepts and
practical tools, such as publishing all building blocks of science in
a research compendium (cf.~\url{https://research-compendium.science/})
or ``repro-pack'' \cite{barba_praxis_2018}.  A completed check means
that code could be executed at least once using the provided
instructions, and, therefore, all code and data was given and could be
investigated more deeply or extended in the future.  Ideally, this is
a ``one~click'' step, but achieving this requires particular skills
and a sufficient level of documentation for third
parties. Furthermore, automation may lead to people gaming the system
or reliance on technology, which can often hide important details.
All such aspects can reduce the understandability of the material, so
we estimate our approach to codechecking, done without automation and
with open human communication, to be a simple way to ensure long-term
transparency and usefulness.  We acknowledge that others have argued
in favour of bitwise reproducibility because, in the long run, it can
be automated (e.g.,
\url{https://twitter.com/khinsen/status/1242842759733665799}), but
until then we need CODECHECK's approach.

\textbf{5. Open by default and transitional by disposition.} \\
Unless there are strong reasons to the contrary
(e.g.,~sensitive data on human subjects), all code and data, both from
author and codechecker, will be made freely available when
the certificate is published.  Openness is not required for the paper
itself, to accommodate journals in their transition to
Open Access models.  The code and data publication should follow
community good practices.  Ultimately we may find that CODECHECKS are
subsumed within peer review.
%% By disposition, a CODECHECK workflow should
%% be designed (a) recognising shortcomings in asserting reproducibility
%% of submissions and (b) with an understanding of the complex situation
%% that led to this. However, establishing a CODECHECK workflow should
%% mirror initiatives to improve education around reproducibility of
%% methods.  The long term goal is for a specific community to reach a
%% state where the level of verification ensured by codecheckers becomes
%% part of peer review and a regular merit of all accepted papers.

@@

We have a curated list of 25 certificates (Table~\ref{tab:register}).
These CODECHECKs fall into three themes: (1) classic and current
papers from computational neuroscience, (2) COVID-19 modelling
preprints, and (3) GIScience.

The first theme was an initial set of papers used to explore the
concept of CODECHECK.  The idea was to take well-known articles from a
domain of interest (Neuroscience).  Our first CODECHECK (certificate
number \texttt{2020-001}) was performed before publication on an
article for the journal \emph{GigaScience}, which visusalized the
outputs from a family of supervised classification algorithms.

The second theme was a response to the COVID-19 pandemic, selecting
papers that predicted outcomes. The checks were solicited through
community interaction or by our initiative rather than requested from
journals.  Some certificates were since acknowledged in the accepted
papers \cite{Davies2020-vj,kucharski_effectiveness_2020}. In
particular, we codechecked the well-known Imperial college model of UK
lockdown procedures from March 2020,
demonstrating that the model results were reproducible
\cite{Chawla2020-hi,cert-2020-010}.

@@

After running the workflow, the codechecker writes a certificate
stating which outputs from the original article, i.e.,~numbers,
figures or tables, could be reproduced.  This certificate is made
openly available so that everyone can see which
elements were reproduced and what limitations or issues were found.
The certificate links to code and data used by the codechecker,
allowing others to build on the work.  The format of the certificates
evolved during the project, as we learnt to automate different
aspects of the certification.  The metadata
is stored in a machine-readable structured file in YAML, the CODECHECK
configuration file \texttt{codecheck.yml}.
The technical specification of the CODECHECK configuration file is
published at \url{https://codecheck.org.uk/spec/config/latest/}.  The
configuration file enables current and future automation of workflows
and meta-analyses.

Figure~\ref{fig:cert} shows pages 1--4 (of 10) of an example
certificate to check predictions of COVID-19 spread across the USA
\cite{cert-2020-012,unwin_report_2020}.  Figure~\ref{fig:cert}A shows
the certificate number and its DOI, which points to the certificate
and any supplemental files on Zenodo.  The CODECHECK logo is added for
recognition and to denote successful reproduction.
Figure~\ref{fig:cert}B provides the key metadata extracted from
\texttt{codecheck.yml}; it names the paper that was checked (title,
DOI), the authors, the codechecker, when the check was performed, and
where code/data are available.
Figure~\ref{fig:cert}C shows a textual summary of how the CODECHECK
was performed and key findings.  Figure~\ref{fig:cert}D
(page 2 of the certificate) shows the outputs that were generated
based on the MANIFEST of output files in the CODECHECK.  It shows the
file name (Output), the description stating to which figure/table each
file should be compared in the original paper (Comment), and the file
size.  Page 3 of the certificate, (Figure~\ref{fig:cert}E) gives
detailed notes from the codechecker, here documenting what steps were
needed to run the code and that the code took about 17~hours to
complete. Finally, page 4 of the certificate shows the first output
generated by the CODECHECK (Figure~\ref{fig:cert}F). In this case, the
figure matched figure~4 of \cite{unwin_report_2020}.  The remaining
pages of the certificate show other outputs and the computing
environment in which the certificate itself was created (not shown
here).

@@

We use freely available infrastructure, GitHub and Zenodo, to run our
system.  The \texttt{codecheckers} \textbf{GitHub} organisation at
\url{https://github.com/codecheckers} contains projects for managing
the project website, the codecheckers community and its discussions,
code repositories, and the main register of CODECHECKs. Both the
project website \url{https://codecheck.org.uk/} and the register at
\url{https://codecheck.org.uk/register} are hosted as GitHub pages.
The register database is a single table in CSV format that connects
the certificate identifier with the repository associated with a
CODECHECK. Each of these repositories, which currently can be hosted
on GitHub or Open Science Framework (OSF), contains the CODECHECK
metadata file \texttt{codecheck.yml}. The register further contains a
column for the type of check, e.g.,~community, journal, or conference,
and the respective GitHub issue where communications and assignments
around a specific check are organised. No information is duplicated
between the register and the metadata files. The continuous
integration infrastructure of GitHub, GitHub Actions, is used to
automate the register for viewing (HTML) and for integration with
other services (CSV, JSON).  \textbf{Zenodo} is our preferred open
repository for storing certificates. It mints DOIs for deposits and
ensures long-term availability of all digital artefacts related to the
project. The CODECHECK community on Zenodo is available at
\url{https://zenodo.org/communities/codecheck/}. It holds
certificates, the regularly archived register
\cite{codecheck_register_jan2021}, and other material related to
CODECHECK.

@@

These tools and resources demonstrate that a CODECHECK process can be
managed without infrastructure costs on freely available platforms.
Automation of some aspects may improve turnaround time. Our main
resource requirements are the \textbf{humans} needed for managing the
project and processes and the codecheckers.  All contributions
currently rely on (partly grant-based) public funding and
volunteering.
